---
title: "An Engineering Manual of Practice for Nature-Based Solutions"
authors: "S. Kyle McKay, Emily Corwin, and Troy Dorman"  
Date: "`r Sys.Date()`"
site: bookdown::bookdown_site
\documentclass: book
output:
  bookdown::bs4_book: default
github-repo: skmckay/NBS.Manual.Of.Practice
---

<span style="color: red;">ALL CONTENT IN THIS REPORT IS PRELIMINARY AND DRAFT. DO NOT REFERENCE FOR ANY PURPOSE.</span>

# {-}

## What is this document? {-}

Infrastructure and environmental managers are increasingly coping with diverse challenges ranging from biodiversity loss and climate change to invasive species and aging infrastructure. These overlapping forms of global change need strategies that embrace ecological outcomes alongside society’s needs for working landscapes. Nature-based solutions (NBS) have emerged as a key pathway for addressing multiple objectives in parallel. NBS, also known as natural or green infrastructure, have have four properties, not all of which may be met by all features : (1) they provide infrastructure functions and services, (2) they consist at least in part of biotic components, (3) they are designed intentionally to achieve environmental, societal, and economic outcomes, and (4) they foster resilience through self-organization.

Designers and engineers implementing infrastructure projects rely on a variety of forms of technical guidance. However, “guidance” spans a huge spectrum of knowledge sharing mechanisms from informal, peer-to-peer learning to formal, international standards. While a variety of interdisciplinary guidelines exist, a lack of engineering design guidance is often cited as slowing the adoption of NBS. 

The American Society of Civil Engineers (ASCE) has initiated development of a manual of practice focused on an engineer’s role in developing and executing NBS projects. In 2025, ASCE partnered with Duke University’s Infrastructure Sustainability and Learning Initiative to develop an outline for a manual of practice (MOP), and in early 2026, the society identified editors and other participants to advance the development of the manual. The general strategy for manual development is to assemble a team and secure funding in 2026, draft content in 2027, and review and publish the manual in 2028. Ultimately, this MOP will provide engineers with the discipline-specific guidance needed to plan, design, and execute projects, establish the professional “standard of care” for NBS, and normalize the use of NBS by municipalities, states, and other infrastructure agencies.

**To be added**:  

- Lay out how this manual fits between interdisciplinary guidance (e.g., International Guidelines for NNBF) and feature-specific guidance (e.g., floodplain benching manuals).  
- Describe the role of subsequent NBS manuals (i.e., Appendices to this document?).  

## Points of contact {-}

The editors of this manual are: 

S. Kyle McKay, Ph.D., P.E.  
Environmental Director, Woolpert  
kyle.mckay at woolpert.com
New York, New York

Emily Corwin, M.S., P.E.  
Interim Executive Director, San Francisco Estuary Institute
emilyc at sfei.org  
Richmond, California

Troy Dorman, Ph.D., P.E., CFM, ENV SP  
Vice President, Director of Water Resources and Sustainability, Halff  
tdorman at halff.com
San Antonio, Texas

The ASCE points of contact are:  

Brian Parsons  
Managing Director, Environmental Water Resources Institute & Chief Sustainability Officer, American Society of Civil Engineers  
bparsons at asce.org  
Washington, D.C.

Ruth Hocker, P.E.  
Director, Technical Community (EWRI/COPRI/CTA), American Society of Civil Engineers  
RHocker at asce.org  
City, State


```{r options, include=FALSE}
#Markdown options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message = FALSE) 

##########
#Clear local memory
rm(list=ls(all=TRUE))

##########
#Load R packages
library(bookdown)
library(downlit)
library(viridis) #Color palette
library(tidyr) #Data wrangling
library(dplyr) #Data manipulation
library(ggplot2) #plotting commands
library(gridExtra) #Arranging figures
library(scales) #visualization stuff
library(kableExtra) #Tables for markdown
library(stringr) #Manipulate strings

##########
#SET COLOR PALETTES
AOC.col <- "#002E5D"
AOC.col1 <- c("#A0291D", "#EB7A28", "#E39D1B", "#FFDB09", "#68954F", "#006955", "#47857D", "#006171", "#5BB6E7", "#4E4A84")  
AOC.col2 <- c("#C2705F", "#F9A35F", "#FCB21B", "#FFEDB2", "#A8CC95", "#96D4CC", "#93BFBA", "#8CC1CD", "#BCE2F5", "#C4C1DF")  
AOC.col3 <- c("#7E6A56", "#552318", "#AA534A", "#61524E", "#EEE1C3", "#E0C5B4", "#D5A897", "#CCC8C4")
```



```{r}
#Example books

#Quintana Go Fish
#Book: https://vmahan1998.github.io/GoFish/
#Github: https://github.com/vmahan1998/GoFish

#NYBEM
#Book: https://mvr-gis.github.io/NYBEM-Report/
#Github: https://github.com/MVR-GIS/NYBEM-Report

#EcoMod training course
#Book: https://usace-wrises.github.io/USACE.EcoMod.Training/
#Github: https://github.com/USACE-WRISES/USACE.EcoMod.Training
```